---
title: "Spatiotemporal CV on gstat's Air Quality Data with xgboost and BART"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatiotemporal CV on gstat's Air Quality Data with xgboost and BART}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

has_deps <- requireNamespace("gstat", quietly = TRUE) &&
  requireNamespace("spacetime", quietly = TRUE) &&
  requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("xgboost", quietly = TRUE) &&
  requireNamespace("dbarts", quietly = TRUE) &&
  requireNamespace("rsample", quietly = TRUE)

knitr::opts_chunk$set(eval = has_deps)
```

```{r, echo = FALSE, eval = !has_deps, results = "asis"}
cat(
  "This vignette needs the **gstat**, **spacetime**, **sf**, **xgboost**,",
  "**dbarts**, and **rsample** packages, which are not all installed.",
  "Install them to render the full vignette."
)
```

## Introduction

This vignette works through a complete `stcvlite` workflow on a real
spatiotemporal dataset: `DE_RB_2005`, the daily rural-background PM10
readings from 69 German monitoring stations in 2005, shipped with
**gstat** (it is the same series used throughout gstat's spatio-temporal
kriging tutorials, originally distributed as `air`/`dates`/`stations` in
**spacetime**). We will

1. reshape the station-day records into the flat `lon`/`lat`/`time` table
   `stcvlite` expects, using the station coordinates and the daily time
   stamps,
2. build spatiotemporal indices with `generate_spt_index()` and inspect
   the spatial blocks that `generate_block_sp_index()` derives from the
   station coordinates,
3. generate every cross-validation scheme `stcvlite` offers and see how
   fold structure differs across them, and
4. fit two very different prediction models -- **xgboost** (gradient
   boosted trees) and **BART** (Bayesian Additive Regression Trees, via
   **dbarts**) -- on the `rset` objects `stcvlite` produces, and compare
   how each cross-validation scheme rates their accuracy.

The point of the last step is not to declare a "winner" model, but to show
*why* the scheme matters: naive random CV lets a model interpolate between
neighboring stations and nearby days, which inflates apparent accuracy.
Leaving out whole stations or whole blocks of time is a stricter, more
realistic test of how these models would perform at a new location or in
a future period.

```{r setup}
library(stcvlite)
library(data.table)
library(sf)
library(rsample)
library(xgboost)
library(dbarts)

set.seed(2026)
```

## 1. The data: gstat's rural-background PM10 stations

`DE_RB_2005` is a `spacetime::STSDF` -- a sparse space-time object holding
one `PM10` measurement per station-day, plus per-station metadata such as
elevation.

```{r load-air}
data("DE_RB_2005", package = "gstat")
DE_RB_2005
```

We flatten it to a long table with `as.data.frame()`, drop missing
readings, and keep the station identifier, its coordinates (originally
UTM zone 32N), the date, station altitude (a genuine predictor of
background PM10), and the response.

```{r flatten}
air_raw <- as.data.frame(DE_RB_2005)
setDT(air_raw)
air_raw <- air_raw[!is.na(PM10)]
air_raw <- air_raw[, .(
  station_id = sp.ID,
  coords.x1, coords.x2,
  time = as.Date(time),
  station_altitude,
  PM10
)]

cat(sprintf(
  "%d station-day observations across %d stations\n",
  nrow(air_raw), uniqueN(air_raw$station_id)
))
```

### Station coordinates

The station coordinates live once per station in `DE_RB_2005@sp`; after
flattening they are repeated on every row. We reproject them from UTM32N
to longitude/latitude (EPSG:4326), matching the convention `stcvlite`
uses for its own sample data (`spdat`), and join them back onto the
long table.

```{r coords}
stations_sf <- sf::st_as_sf(
  unique(air_raw[, .(station_id, coords.x1, coords.x2)]),
  coords = c("coords.x1", "coords.x2"), crs = 32632
)
stations_sf <- sf::st_transform(stations_sf, 4326)
station_coords <- data.table(
  station_id = stations_sf$station_id,
  lon = sf::st_coordinates(stations_sf)[, 1],
  lat = sf::st_coordinates(stations_sf)[, 2]
)

air_dt <- merge(air_raw, station_coords, by = "station_id")
```

We add a pair of day-of-year harmonics so the prediction models have a
smooth seasonal signal to work with, alongside the station coordinates
and altitude.

```{r features}
air_dt[, doy := as.integer(format(time, "%j"))]
air_dt[, doy_sin := sin(2 * pi * doy / 365)]
air_dt[, doy_cos := cos(2 * pi * doy / 365)]
air_dt[, c("coords.x1", "coords.x2", "doy") := NULL]
setcolorder(air_dt, c("station_id", "lon", "lat", "time"))

str(air_dt)
```

## 2. Preparing the covariate table

`stcvlite`'s index generators expect a plain `data.frame`-like object with
`lon`, `lat`, and `time` columns (plus whatever covariates you like) --
no special wrapper class. `prep_input()` does the coordinate/time
extraction for `sf`, `sftime`, `SpatVector`, and `SpatRasterDataset`
inputs and returns exactly that shape, so we hand it an `sf` version of
`air_dt` and use its output as-is.

```{r stdt}
air_sf <- sf::st_as_sf(air_dt, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
air_prepped <- suppressWarnings(stcvlite::prep_input(air_sf))

class(air_prepped)
air_prepped
```

(`prep_input()` calls `sf::st_point_on_surface()` internally, which warns
for point geometries even though the coordinates it returns are exact --
safe to ignore here.)

## 3. Spatiotemporal indices and station-coordinate blocks

`generate_spt_index()` turns the station coordinates (and, optionally,
time) into a single grouping key. In `"spatial"` mode it collapses to one
key per station; in `"spatiotemporal"` mode every station-day gets its
own key.

```{r spt-index}
air_spatial <- generate_spt_index(air_prepped, mode = "spatial")
air_spt <- generate_spt_index(air_prepped, mode = "spatiotemporal")

cat(sprintf(
  "unique spatial keys: %d (= number of stations)\n",
  uniqueN(air_spatial$sp_index)
))
cat(sprintf(
  "unique spatiotemporal keys: %d (= number of rows)\n",
  uniqueN(air_spt$sp_index)
))
```

`generate_block_sp_index()` uses the same `lon`/`lat` columns to group
stations into contiguous spatial blocks (DBSCAN by default), clustering
each *unique* station location once regardless of how many days it was
observed. This is the mechanism behind the `"lblo"` and `"lblto"` schemes
below -- six blocks of stations, chosen purely from their coordinates:

```{r block-index}
air_blocks <- generate_block_sp_index(air_prepped, cv_fold = 6)
table(air_blocks$sp_index)
```

```{r block-plot, eval = has_deps && requireNamespace("plotly", quietly = TRUE)}
plot_cv_folds(air_blocks, air_blocks$sp_index, max_points = 5000)
```

## 4. Every cross-validation scheme, side by side

`generate_cv_index()` implements seven schemes (see the package README
for the full description of each acronym). We build all of them here to
compare the fold structure they induce on the same 23,230-row dataset:
leave-one-station-out (`lolo`) and leave-one-day-out (`loto`) each hold
out one coordinate at a time; `lolto` is plain leave-one-row-out;
`lblo`/`lbto`/`lblto` group stations and/or days into contiguous blocks;
and `random` is the naive baseline.

```{r all-modes}
cv_indices <- list(
  lolo   = generate_cv_index(air_prepped, cv_mode = "lolo"),
  loto   = generate_cv_index(air_prepped, cv_mode = "loto"),
  lolto  = generate_cv_index(air_prepped, cv_mode = "lolto"),
  lblo   = generate_cv_index(air_prepped, cv_mode = "lblo", cv_fold = 6),
  lbto   = generate_cv_index(air_prepped, cv_mode = "lbto", cv_fold = 6),
  lblto  = generate_cv_index(air_prepped, cv_mode = "lblto", sp_fold = 4, t_fold = 3),
  random = generate_cv_index(air_prepped, cv_mode = "random", cv_fold = 5)
)

fold_summary <- rbindlist(lapply(names(cv_indices), function(m) {
  tb <- table(cv_indices[[m]])
  data.table(
    cv_mode = m,
    n_folds = length(tb),
    min_fold_size = min(tb),
    max_fold_size = max(tb)
  )
}))
knitr::kable(fold_summary)
```

`lolo` and `lblo`/`lbto`/`lblto` hold out *whole stations or blocks*, so
every fold tests extrapolation in space and/or time. `loto` and `lolto`
produce far more, much smaller folds (365 and 23,230 respectively) --
faithful to the "leave-one-out" definition, but too many to refit two
models on for this vignette. We generate their indices to show the fold
structure, and evaluate models on the other five schemes below, using
the same helper code that would apply identically to `loto`/`lolto` given
enough time.

## 5. Prediction models: xgboost and BART

Both models use the same feature set -- station coordinates, altitude,
and the seasonal harmonics -- and are refit independently on the
analysis set of every fold, then scored on that fold's assessment set.

```{r model-helpers}
feature_cols <- c("lon", "lat", "station_altitude", "doy_sin", "doy_cos")
target_col <- "PM10"

fit_predict_xgb <- function(train, test) {
  dtrain <- xgboost::xgb.DMatrix(
    data = as.matrix(train[, ..feature_cols]),
    label = train[[target_col]]
  )
  fit <- xgboost::xgb.train(
    data = dtrain,
    nrounds = 150,
    params = list(
      objective = "reg:squarederror",
      max_depth = 4, eta = 0.1, subsample = 0.8
    ),
    verbose = 0
  )
  predict(fit, as.matrix(test[, ..feature_cols]))
}

fit_predict_bart <- function(train, test) {
  fit <- dbarts::bart(
    x.train = as.matrix(train[, ..feature_cols]),
    y.train = train[[target_col]],
    x.test = as.matrix(test[, ..feature_cols]),
    ntree = 50, ndpost = 300, nskip = 100,
    verbose = FALSE, keeptrees = FALSE
  )
  fit$yhat.test.mean
}
```

`convert_cv_index_rset()` turns a fold-assignment vector into a
`tidymodels`-compatible `rset`; `rsample::analysis()` and
`rsample::assessment()` pull the train/test tables out of each split.
`evaluate_rset()` loops over every fold of a given `rset`, fits both
models, and records RMSE/MAE:

```{r evaluate-helper}
rmse <- function(actual, pred) sqrt(mean((actual - pred)^2))
mae <- function(actual, pred) mean(abs(actual - pred))

evaluate_rset <- function(rset, cv_mode) {
  per_fold <- lapply(seq_len(nrow(rset)), function(i) {
    split <- rset$splits[[i]]
    train <- as.data.table(rsample::analysis(split))
    test <- as.data.table(rsample::assessment(split))

    pred_xgb <- fit_predict_xgb(train, test)
    pred_bart <- fit_predict_bart(train, test)

    data.table(
      cv_mode = cv_mode,
      fold = rset$id[i],
      n_test = nrow(test),
      engine = c("xgboost", "bart"),
      rmse = c(rmse(test[[target_col]], pred_xgb), rmse(test[[target_col]], pred_bart)),
      mae = c(mae(test[[target_col]], pred_xgb), mae(test[[target_col]], pred_bart))
    )
  })
  rbindlist(per_fold)
}
```

We run this on five schemes: `random` as the (optimistic) baseline,
`lolo` for pure spatial extrapolation to unseen stations, `lblo` for
block-wise spatial extrapolation, `lbto` for pure temporal extrapolation
to unseen time blocks, and `lblto` for combined spatiotemporal block
extrapolation. `lolo` alone means refitting both models 69 times; on a
laptop this section takes roughly two to three minutes.

```{r run-eval}
rset_random <- convert_cv_index_rset(cv_indices$random, air_prepped, cv_mode = "random")
rset_lolo   <- convert_cv_index_rset(cv_indices$lolo, air_prepped, cv_mode = "lolo")
rset_lblo   <- convert_cv_index_rset(cv_indices$lblo, air_prepped, cv_mode = "lblo")
rset_lbto   <- convert_cv_index_rset(cv_indices$lbto, air_prepped, cv_mode = "lbto")
rset_lblto  <- convert_cv_index_rset(cv_indices$lblto, air_prepped, cv_mode = "lblto")

results <- rbindlist(list(
  evaluate_rset(rset_random, "random"),
  evaluate_rset(rset_lolo, "lolo"),
  evaluate_rset(rset_lblo, "lblo"),
  evaluate_rset(rset_lbto, "lbto"),
  evaluate_rset(rset_lblto, "lblto")
))
```

## 6. Comparing the schemes

Averaging RMSE within each scheme/engine combination shows how much the
CV scheme itself changes the apparent accuracy:

```{r summary-table}
summary_tbl <- results[, .(
  n_folds = uniqueN(fold),
  mean_rmse = mean(rmse),
  mean_mae = mean(mae)
), by = .(cv_mode, engine)]

knitr::kable(summary_tbl[order(cv_mode, engine)], digits = 2)
```

```{r summary-plot, eval = has_deps && requireNamespace("plotly", quietly = TRUE)}
plotly::plot_ly(
  data = results,
  x = ~cv_mode, y = ~rmse, color = ~engine,
  type = "box"
) |>
  plotly::layout(
    yaxis = list(title = "Fold RMSE (PM10, ug/m3)"),
    xaxis = list(title = "CV scheme"),
    boxmode = "group"
  )
```

## Takeaways

* `random` CV lets both models lean on neighboring stations and adjacent
  days, so it reads as the easiest scheme -- and it is: mean RMSE for
  both engines sits lowest here.
* `lolo` and `lblo` remove entire stations from training. RMSE rises only
  modestly over `random`, which makes sense given the feature set:
  longitude, latitude, and altitude are smooth surfaces, so a model can
  extrapolate them reasonably well to a station it has never seen.
* `lbto` removes contiguous *blocks of days* instead, and RMSE roughly
  doubles relative to `random` for both engines. The seasonal harmonics
  (`doy_sin`/`doy_cos`) repeat every year, so once a block of consecutive
  days is entirely held out, neither model has anything in the feature
  set that encodes the actual meteorology of that stretch of time -- only
  its position in the yearly cycle. `lblto` combines both stresses and
  lands in between.
* xgboost and BART track each other closely within every scheme; the gap
  between schemes is driven far more by *what* is held out than by
  *which* model is used. That gap -- not the model choice -- is what
  `stcvlite`'s CV schemes are built to expose, and it is invisible if you
  only ever run random k-fold.

The `rset` objects used above (`rset_random`, `rset_lolo`, ...) are
ordinary `rsample` objects, so they also drop directly into
`parsnip`/`workflows`/`tune` if you prefer that interface -- see the
README's "Integration with tidymodels" section for the `fit_resamples()`
form of the same workflow.
