---
title: "Spatiotemporal CV on Air Quality Data with torch Neural Networks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatiotemporal CV on Air Quality Data with torch Neural Networks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)

has_deps <- requireNamespace("gstat", quietly = TRUE) &&
  requireNamespace("spacetime", quietly = TRUE) &&
  requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("torch", quietly = TRUE) &&
  requireNamespace("rsample", quietly = TRUE)

```

## Note
Please note that this vignette is intended for demonstration with CUDA-enabled GPUs. The vignette is intentionally not run to avoid long build times and continuous integration issues.

## Introduction

This is a companion to the
[`air-xgboost-bart`](air-xgboost-bart.html) vignette. It uses the same
dataset -- `DE_RB_2005`, daily rural-background PM10 readings from 69
German monitoring stations in 2005, shipped with **gstat** -- but fits a
small feed-forward neural network with **torch** instead of tree-based
models, and folds it into `stcvlite`'s spatiotemporal CV schemes the same
way.

Neural networks bring two practical concerns that tree ensembles mostly
sidestep: features must be scaled (gradient descent on raw longitude/PM10
scales does not converge well), and training is iterative, so we need a
loop over epochs and mini-batches rather than a single `fit()` call. This
vignette focuses on that machinery -- scaling *within* each fold to avoid
leakage, wiring a `torch::dataset`/`dataloader` to `stcvlite`'s `rset`
splits, and reading a training curve -- rather than re-deriving the data
preparation steps, which the companion vignette covers in detail.

The vignette has two parts. Part 1 (Sections 1-7) fits a plain
feed-forward network (an MLP) row by row, exactly like the xgboost/BART
models in the companion vignette -- every station-day is an independent
observation, and the network never sees which station or day a row came
from except through its `lon`/`lat`/time-harmonic features. Part 2
(Sections 8-12) fits a **graph neural network** (GNN) instead: the 69
stations become nodes of a fixed spatial graph, and each layer lets a
station's prediction draw on its geographic neighbors' features directly,
which is a more natural match for `stcvlite`'s station-coordinate CV
schemes (`lolo`, `lblo`, ...) than treating every row as independent.

```{r setup}
library(stcvlite)
library(data.table)
library(sf)
library(rsample)
library(torch)

set.seed(2026)
torch::torch_manual_seed(2026)

device <- if (torch::cuda_is_available()) {
  torch::torch_device("cuda")
} else if (torch::backends_mps_is_available()) {
  torch::torch_device("mps")
} else {
  torch::torch_device("cpu")
}
device
```

## 1. Data and features

Flatten `DE_RB_2005`, reproject the station coordinates to lon/lat, and
add day-of-year harmonics -- identical to the companion vignette.

```{r load-air}
data("DE_RB_2005", package = "gstat")

air_raw <- as.data.frame(DE_RB_2005)
setDT(air_raw)
air_raw <- air_raw[!is.na(PM10)]
air_raw <- air_raw[, .(
  station_id = sp.ID,
  coords.x1, coords.x2,
  time = as.Date(time),
  station_altitude,
  PM10
)]

stations_sf <- sf::st_as_sf(
  unique(air_raw[, .(station_id, coords.x1, coords.x2)]),
  coords = c("coords.x1", "coords.x2"), crs = 32632
)
stations_sf <- sf::st_transform(stations_sf, 4326)
station_coords <- data.table(
  station_id = stations_sf$station_id,
  lon = sf::st_coordinates(stations_sf)[, 1],
  lat = sf::st_coordinates(stations_sf)[, 2]
)

air_dt <- merge(air_raw, station_coords, by = "station_id")
air_dt[, doy := as.integer(format(time, "%j"))]
air_dt[, doy_sin := sin(2 * pi * doy / 365)]
air_dt[, doy_cos := cos(2 * pi * doy / 365)]
air_dt[, c("coords.x1", "coords.x2", "doy") := NULL]
setcolorder(air_dt, c("station_id", "lon", "lat", "time"))

feature_cols <- c("lon", "lat", "station_altitude", "doy_sin", "doy_cos")
target_col <- "PM10"

nrow(air_dt)
```

```{r prep}
air_sf <- sf::st_as_sf(air_dt, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
air_prepped <- suppressWarnings(stcvlite::prep_input(air_sf))
```

## 2. Cross-validation indices

We compare four schemes: `random` as the optimistic baseline, `lblo` and
`lbto` for block-wise spatial and temporal extrapolation, and `lblto` for
both combined. Unlike the companion vignette, we skip `lolo`/`loto`/
`lolto`: those hold out one station or one day at a time, which means 69
or 365 full network retrainings, and unlike xgboost/BART, retraining a
network is not cheap enough to do that many times in a vignette.

```{r cv-indices}
cv_indices <- list(
  random = generate_cv_index(air_prepped, cv_mode = "random", cv_fold = 5),
  lblo   = generate_cv_index(air_prepped, cv_mode = "lblo", cv_fold = 6),
  lbto   = generate_cv_index(air_prepped, cv_mode = "lbto", cv_fold = 6),
  lblto  = generate_cv_index(air_prepped, cv_mode = "lblto", sp_fold = 3, t_fold = 2)
)

rset_list <- Map(
  convert_cv_index_rset,
  cvindex = cv_indices,
  cv_mode = names(cv_indices),
  MoreArgs = list(data = air_prepped)
)
sapply(rset_list, nrow)
```

## 3. Scaling within each fold

Gradient descent needs features on comparable scales, and the target
benefits from centering too. Fitting the scaler on the *analysis*
(training) set only -- and applying those same parameters to the
*assessment* (test) set -- keeps test-fold information out of training,
exactly as a model itself must never see assessment rows.

```{r scale-helper}
scale_with <- function(x, center, scale) {
  scale <- ifelse(scale == 0, 1, scale)
  sweep(sweep(x, 2, center, "-"), 2, scale, "/")
}

prep_fold_matrices <- function(train, test, feature_cols, target_col) {
  x_train_raw <- as.matrix(train[, ..feature_cols])
  x_test_raw <- as.matrix(test[, ..feature_cols])

  x_center <- colMeans(x_train_raw)
  x_scale <- apply(x_train_raw, 2, sd)
  y_center <- mean(train[[target_col]])
  y_scale <- sd(train[[target_col]])

  list(
    x_train = scale_with(x_train_raw, x_center, x_scale),
    x_test = scale_with(x_test_raw, x_center, x_scale),
    y_train = (train[[target_col]] - y_center) / y_scale,
    y_center = y_center,
    y_scale = y_scale
  )
}
```

We scale the response too, purely so the loss starts in a numerically
sane range for the optimizer; predictions are un-scaled back to
micrograms per cubic meter before computing RMSE/MAE.

## 4. The network

A small two-hidden-layer MLP is enough for five numeric features. `torch`
models are declared with `nn_module()`: `initialize()` creates the
layers, `forward()` wires them together.

```{r model-def}
pm10_net <- torch::nn_module(
  classname = "pm10_net",
  initialize = function(n_features, hidden = c(16, 8), dropout = 0.1) {
    self$fc1 <- torch::nn_linear(n_features, hidden[1])
    self$fc2 <- torch::nn_linear(hidden[1], hidden[2])
    self$fc3 <- torch::nn_linear(hidden[2], 1)
    self$drop <- torch::nn_dropout(dropout)
  },
  forward = function(x) {
    x |>
      self$fc1() |>
      torch::nnf_relu() |>
      self$drop() |>
      self$fc2() |>
      torch::nnf_relu() |>
      self$fc3()
  }
)
```

`dropout` is applied only when the module is in training mode
(`model$train()`); calling `model$eval()` before prediction switches it
off, which matters below.

A `torch::dataset` wraps the scaled matrices so `dataloader()` can
shuffle and batch them for us:

```{r dataset-def}
pm10_dataset <- torch::dataset(
  name = "pm10_dataset",
  initialize = function(x, y) {
    self$x <- torch::torch_tensor(x, dtype = torch::torch_float())
    self$y <- torch::torch_tensor(matrix(y, ncol = 1), dtype = torch::torch_float())
  },
  .getitem = function(i) {
    list(x = self$x[i, ], y = self$y[i, ])
  },
  .length = function() {
    self$x$size(1)
  }
)
```

## 5. Training and prediction helper

`fit_predict_torch()` scales the fold, trains for a fixed number of
epochs with Adam and MSE loss, and returns predictions on the assessment
set (un-scaled) plus the per-epoch training loss for diagnostics. To keep
this vignette's build time reasonable, it also caps the *analysis*
(training) rows it fits on per fold at `max_train`, subsampling down from
however many the fold actually has; the full assessment set is still
scored either way, since prediction is a single forward pass and costs
essentially nothing regardless of size. A smaller sample is also a
reasonable match for the smaller network (`hidden = c(16, 8)`) defined
above -- there is less point training a wider net on a subsample too
small to need that much capacity.

```{r train-helper}
fit_predict_torch <- function(train, test, feature_cols, target_col,
                               epochs = 15, batch_size = 512, lr = 1e-3,
                               max_train = 4000) {
  if (nrow(train) > max_train) {
    train <- train[sample.int(nrow(train), max_train)]
  }
  fold <- prep_fold_matrices(train, test, feature_cols, target_col)

  model <- pm10_net(n_features = length(feature_cols))
  model <- model$to(device = device)
  optimizer <- torch::optim_adam(model$parameters, lr = lr)

  train_ds <- pm10_dataset(fold$x_train, fold$y_train)
  train_dl <- torch::dataloader(train_ds, batch_size = batch_size, shuffle = TRUE)

  losses <- numeric(epochs)
  for (epoch in seq_len(epochs)) {
    model$train()
    epoch_loss <- 0
    n_seen <- 0
    coro::loop(for (batch in train_dl) {
      x_b <- batch$x$to(device = device)
      y_b <- batch$y$to(device = device)

      optimizer$zero_grad()
      pred <- model(x_b)
      loss <- torch::nnf_mse_loss(pred, y_b)
      loss$backward()
      optimizer$step()

      epoch_loss <- epoch_loss + loss$item() * x_b$size(1)
      n_seen <- n_seen + x_b$size(1)
    })
    losses[epoch] <- epoch_loss / n_seen
  }

  model$eval()
  x_test_t <- torch::torch_tensor(fold$x_test, dtype = torch::torch_float())$to(device = device)
  pred_scaled <- torch::with_no_grad(model(x_test_t))
  pred <- as.numeric(pred_scaled$to(device = "cpu")) * fold$y_scale + fold$y_center

  list(pred = pred, loss_curve = losses)
}
```

Three details carry over directly from `dbarts`/`xgboost` fold-fitting
in the companion vignette, just implemented differently: refit from
scratch on every fold's analysis set (a fresh `pm10_net()` each call, so
no weights leak across folds), score only on that fold's assessment set,
and record both RMSE and MAE.

```{r evaluate-helper}
rmse <- function(actual, pred) sqrt(mean((actual - pred)^2))
mae <- function(actual, pred) mean(abs(actual - pred))

evaluate_rset_torch <- function(rset, cv_mode, epochs = 15) {
  per_fold <- lapply(seq_len(nrow(rset)), function(i) {
    split <- rset$splits[[i]]
    train <- as.data.table(rsample::analysis(split))
    test <- as.data.table(rsample::assessment(split))

    fit <- fit_predict_torch(train, test, feature_cols, target_col, epochs = epochs)

    data.table(
      cv_mode = cv_mode,
      fold = rset$id[i],
      n_test = nrow(test),
      rmse = rmse(test[[target_col]], fit$pred),
      mae = mae(test[[target_col]], fit$pred),
      final_train_loss = fit$loss_curve[epochs]
    )
  })
  rbindlist(per_fold)
}
```

## 6. Running the four schemes

This refits the network `r sum(sapply(rset_list, nrow))` times in total
(5 + 6 + 6 + 6 folds). Each fit trains on at most `max_train = 4000`
subsampled rows for 15 epochs in mini-batches of 512, which keeps this
section under two minutes on a laptop CPU. That is a real
speed/precision trade-off, not a free lunch: fewer training rows and
fewer epochs mean noisier per-fold RMSE than a full-data, longer-trained
run would give. Raise `max_train` (up to `Inf`, i.e. no subsampling) and
`epochs` if you want tighter numbers and can spare on the order of ten
extra minutes of build time. Section 12 revisits mini-batch granularity
for the graph model, where the relevant batch unit is a *day*, not a
row, and no subsampling is needed to stay fast.

```{r run-eval}
results <- rbindlist(Map(evaluate_rset_torch, rset_list, names(rset_list)))

summary_tbl <- results[, .(
  n_folds = uniqueN(fold),
  mean_rmse = mean(rmse),
  mean_mae = mean(mae)
), by = cv_mode]

knitr::kable(summary_tbl[order(cv_mode)], digits = 2)
```

```{r summary-plot, eval = has_deps && requireNamespace("plotly", quietly = TRUE)}
plotly::plot_ly(
  data = results,
  x = ~cv_mode, y = ~rmse,
  type = "box"
) |>
  plotly::layout(
    yaxis = list(title = "Fold RMSE (PM10, ug/m3)"),
    xaxis = list(title = "CV scheme")
  )
```

## 7. Reading a training curve

Aggregate RMSE hides whether a given fold's network actually converged.
Re-fitting one `lblo` fold and plotting its loss curve shows what a
healthy run looks like -- useful as a sanity check before trusting the
numbers above, and as a template for spotting an under- or over-fit fold
(a curve that is still falling at the last epoch means `epochs` was set
too low; one that is flat from the start usually means `lr` is too low
or too high).

```{r loss-curve}
split1 <- rset_list$lblo$splits[[1]]
train1 <- as.data.table(rsample::analysis(split1))
test1 <- as.data.table(rsample::assessment(split1))

fit1 <- fit_predict_torch(train1, test1, feature_cols, target_col)

plot(
  seq_along(fit1$loss_curve), fit1$loss_curve,
  type = "l", xlab = "epoch", ylab = "training MSE (scaled PM10)",
  main = "Training curve, lblo fold 1"
)
```

## MLP takeaways

* Neural networks need scaling; tree ensembles do not. Fitting the
  scaler on the analysis set alone, per fold, is the same leakage
  discipline as any other model -- the difference is that with trees
  there is nothing to leak (splits are scale-invariant), so it is easy to
  forget this step when switching to `torch`.
* `torch::dataset()` + `dataloader()` slots directly into the
  `rsample::analysis()`/`assessment()` tables that `convert_cv_index_rset()`
  produces -- the CV machinery in `stcvlite` does not care what kind of
  model consumes its splits.
* Refitting cost is the practical constraint on which schemes you can
  afford to run: `lolo`'s 69 stations were fine for xgboost/BART in the
  companion vignette but would mean 69 separate training loops here.
  Block schemes (`lblo`/`lbto`/`lblto`) keep the fold count -- and
  therefore the number of networks trained -- small regardless of
  dataset size, which makes them a more practical default for
  iterative models.
* As in the companion vignette, expect `lbto`/`lblto` (temporal
  extrapolation) to read as harder than `random` or `lblo`, for the same
  reason: the seasonal harmonics repeat every year, so a network that
  has never seen a given stretch of days has no signal about that
  stretch's actual meteorology, only its position in the annual cycle.

## 8. From independent rows to a graph

The MLP above scores every station-day row on its own; if a station is
held out entirely (`lolo`/`lblo`), the network's only handle on it is
whatever it can extrapolate from `lon`/`lat`/`station_altitude` as smooth
functions. A **graph neural network** gives the model something more
direct: an explicit edge to each held-out station's nearest neighbors, so
its prediction is built partly from *their* observed behavior, not just
from a coordinate-to-PM10 function learned elsewhere on the map.

This section builds a small graph convolutional model over the 69
station locations and reuses the exact `cv_indices` computed in
Section 2 -- the fold assignments themselves do not change; only the
model, and the shape of data it expects, do.

## 9. Building the station graph

Nodes are the 69 unique stations. Edges connect each station to its
`k` nearest neighbors by coordinate distance, found with `FNN::get.knn()`
(the same package `stcvlite` uses internally for one of its own spatial
blocking engines). We symmetrize the result -- if A lists B as a
neighbor, we keep the A-B edge even if B's own `k` nearest happen not to
include A -- and store a **row-normalized** adjacency matrix, so each
node's graph-convolution input is the *mean* of its neighbors' features.

```{r graph-build}
station_lookup <- unique(air_dt[, .(station_id, lon, lat, station_altitude)])
setorder(station_lookup, station_id)
station_lookup[, station_idx := .I]
n_stations <- nrow(station_lookup)

knn <- FNN::get.knn(as.matrix(station_lookup[, .(lon, lat)]), k = 8)
adj <- matrix(0, n_stations, n_stations)
for (i in seq_len(n_stations)) adj[i, knn$nn.index[i, ]] <- 1
adj <- pmax(adj, t(adj))

a_mean <- adj / rowSums(adj)
a_mean_t <- torch::torch_tensor(a_mean, dtype = torch::torch_float())$to(device = device)

cat(sprintf(
  "%d stations, %d undirected edges (k = 8 nearest neighbors)\n",
  n_stations, sum(adj) / 2
))
```

A first version of this section used a plain symmetric-normalized
adjacency (the shared-weight aggregation from Kipf & Welling's original
GCN paper, `D^-1/2 (A + I) D^-1/2`). On this graph it consistently
underperformed the row-level MLP: mixing a station's own signal in with
eight neighbors under one shared weight matrix diluted exactly the
per-station information (altitude, precise location) that mattered most.
The layer below fixes that by giving a station's own features and its
neighbors' features **separate** weights -- a GraphSAGE-style mean
aggregator -- which recovered MLP-competitive accuracy in testing.

## 10. From rows to a station-day panel

A row-level `x`/`y` matrix does not have a graph structure to convolve
over. The graph convolution needs, for each *day*, a feature matrix with
one row per station in a fixed order -- so we reshape the data into a
`(day x station x feature)` array instead. Not every station reports
every day (54-68 of the 69 report on any given day), so we track which
`(day, station)` cells are actually observed with a boolean mask.

```{r panel-build}
day_lookup <- data.table(time = sort(unique(air_dt$time)))
day_lookup[, day_idx := .I]
day_lookup[, doy := as.integer(format(time, "%j"))]
day_lookup[, doy_sin := sin(2 * pi * doy / 365)]
day_lookup[, doy_cos := cos(2 * pi * doy / 365)]
n_days <- nrow(day_lookup)

air_panel <- copy(air_dt)
air_panel[, station_idx := station_lookup$station_idx[match(station_id, station_lookup$station_id)]]
air_panel[, day_idx := day_lookup$day_idx[match(time, day_lookup$time)]]

range(air_panel[, .N, by = day_idx]$N)
```

`station_altitude`/`lon`/`lat` are known for every station regardless of
whether it reported PM10 on a given day, and `doy_sin`/`doy_cos` are
known for every calendar day regardless of station -- so building the
full `(day, station)` feature grid does not invent any information, it
only reshapes rows we already had. Only the *target* (`PM10`) and the
*fold assignment* are actually missing outside the observed cells, and
both are masked out below rather than filled in.

```{r feature-array}
station_feat <- scale(as.matrix(station_lookup[order(station_idx), .(lon, lat, station_altitude)]))
day_feat <- as.matrix(day_lookup[order(day_idx), .(doy_sin, doy_cos)])

x_station <- torch::torch_tensor(station_feat, dtype = torch::torch_float())
x_day <- torch::torch_tensor(day_feat, dtype = torch::torch_float())

x_station_b <- x_station$unsqueeze(1)$expand(c(n_days, n_stations, ncol(station_feat)))
x_day_b <- x_day$unsqueeze(2)$expand(c(n_days, n_stations, ncol(day_feat)))
x_full <- torch::torch_cat(list(x_station_b, x_day_b), dim = 3)

x_full$shape
```

```{r mask-build}
y_full <- matrix(0, n_days, n_stations)
obs_mask <- matrix(FALSE, n_days, n_stations)
idx_mat <- as.matrix(air_panel[, .(day_idx, station_idx)])
y_full[idx_mat] <- air_panel[[target_col]]
obs_mask[idx_mat] <- TRUE

fold_mats <- lapply(cv_indices, function(cv) {
  m <- matrix(NA_integer_, n_days, n_stations)
  m[idx_mat] <- cv
  m
})
```

`station_feat`/`day_feat` are standardized once, globally, rather than
per fold: unlike the MLP's target, these are exogenous covariates
(station coordinates, altitude, calendar date) that do not depend on the
response, so there is no leakage risk in using every station's and every
day's value to compute the mean/sd. The target is still scaled per fold,
from the training mask only, exactly as in Section 3.

## 11. A GraphSAGE-style model

Each layer computes two linear transforms -- one of the node's own
features, one of its neighbors' mean features -- and adds them, rather
than mixing both under one shared weight matrix:

```{r sage-model}
sage_layer <- torch::nn_module(
  classname = "sage_layer",
  initialize = function(in_features, out_features, a_norm) {
    self$lin_self <- torch::nn_linear(in_features, out_features)
    self$lin_neigh <- torch::nn_linear(in_features, out_features)
    self$register_buffer("a_norm", a_norm)
  },
  forward = function(x) {
    neighbor_mean <- torch::torch_matmul(self$a_norm, x)
    self$lin_self(x) + self$lin_neigh(neighbor_mean)
  }
)

pm10_sage <- torch::nn_module(
  classname = "pm10_sage",
  initialize = function(n_features, a_norm, hidden = c(32, 16), dropout = 0.05) {
    self$gc1 <- sage_layer(n_features, hidden[1], a_norm)
    self$gc2 <- sage_layer(hidden[1], hidden[2], a_norm)
    self$out <- torch::nn_linear(hidden[2], 1)
    self$drop <- torch::nn_dropout(dropout)
  },
  forward = function(x) {
    x |>
      self$gc1() |>
      torch::nnf_relu() |>
      self$drop() |>
      self$gc2() |>
      torch::nnf_relu() |>
      self$out()
  }
)
```

`self$register_buffer("a_norm", a_norm)` matters: the adjacency matrix is
fixed (not learned), but registering it as a buffer -- rather than just
assigning `self$a_norm <- a_norm` -- makes `model$to(device = ...)` move
it along with the learnable weights, so it stays on the same device as
everything else if you switch to a GPU.

Because `a_norm` is a single `(69 x 69)` matrix shared by every day,
`torch_matmul(a_norm, x)` broadcasts over a whole batch of days at once
when `x` has shape `(batch, 69, features)` -- one call aggregates
neighbor information for every day in the batch, no per-day loop needed.

## 12. Training across folds

A "batch" here is a set of *days* rather than a set of rows: training
loops over shuffled day indices, and for each batch pulls
`x_full[days, , ]`, forwards it through the shared graph, and computes a
masked MSE that only counts `(day, station)` cells the current fold
assigns to the analysis set. Days that belong entirely to the assessment
set for a given fold (e.g. every day in a held-out `lbto` block) are
skipped during training and used only for scoring.

```{r gcn-train-helper}
masked_mse <- function(pred, target, mask) {
  ((pred - target)^2 * mask)$sum() / mask$sum()
}

fit_predict_gcn <- function(cv_mode, fold_id, epochs = 80, batch_days = 32, lr = 0.01) {
  fold_mat <- fold_mats[[cv_mode]]
  train_mask_mat <- obs_mask & !is.na(fold_mat) & (fold_mat != fold_id)
  test_mask_mat  <- obs_mask & !is.na(fold_mat) & (fold_mat == fold_id)

  y_center <- mean(y_full[train_mask_mat])
  y_scale <- sd(y_full[train_mask_mat])
  y_scaled <- (y_full - y_center) / y_scale
  y_scaled[!obs_mask] <- 0

  y_t <- torch::torch_tensor(y_scaled, dtype = torch::torch_float())$to(device = device)
  train_mask_t <- torch::torch_tensor(train_mask_mat * 1, dtype = torch::torch_float())$to(device = device)

  train_days <- which(rowSums(train_mask_mat) > 0)
  test_days <- which(rowSums(test_mask_mat) > 0)

  model <- pm10_sage(n_features = x_full$shape[3], a_norm = a_mean_t)
  model <- model$to(device = device)
  optimizer <- torch::optim_adam(model$parameters, lr = lr)

  losses <- numeric(epochs)
  for (epoch in seq_len(epochs)) {
    model$train()
    perm <- sample(train_days)
    batches <- split(perm, ceiling(seq_along(perm) / batch_days))
    epoch_loss <- 0
    for (b in batches) {
      optimizer$zero_grad()
      x_b <- x_full[b, , ]$to(device = device)
      if (length(b) == 1) x_b <- x_b$unsqueeze(1)
      pred <- model(x_b)$squeeze(3)
      y_b <- y_t[b, ]
      m_b <- train_mask_t[b, ]
      if (length(b) == 1) {
        y_b <- y_b$unsqueeze(1)
        m_b <- m_b$unsqueeze(1)
      }
      loss <- masked_mse(pred, y_b, m_b)
      loss$backward()
      optimizer$step()
      epoch_loss <- epoch_loss + loss$item() * length(b)
    }
    losses[epoch] <- epoch_loss / length(perm)
  }

  model$eval()
  x_test <- x_full[test_days, , ]$to(device = device)
  if (length(test_days) == 1) x_test <- x_test$unsqueeze(1)
  pred_scaled <- torch::with_no_grad(model(x_test)$squeeze(3))
  pred_mat <- as.matrix(pred_scaled$to(device = "cpu")) * y_scale + y_center

  actual_mat <- y_full[test_days, , drop = FALSE]
  mask_mat <- test_mask_mat[test_days, , drop = FALSE]

  list(
    pred = pred_mat[mask_mat == 1],
    actual = actual_mat[mask_mat == 1],
    loss_curve = losses
  )
}

evaluate_gcn <- function(cv_mode, epochs = 80) {
  n_folds <- max(cv_indices[[cv_mode]])
  per_fold <- lapply(seq_len(n_folds), function(k) {
    fit <- fit_predict_gcn(cv_mode, k, epochs = epochs)
    data.table(
      cv_mode = cv_mode,
      fold = k,
      n_test = length(fit$pred),
      rmse = rmse(fit$actual, fit$pred),
      mae = mae(fit$actual, fit$pred)
    )
  })
  rbindlist(per_fold)
}
```

```{r gcn-run-eval}
results_gcn <- rbindlist(lapply(names(cv_indices), evaluate_gcn))

summary_gcn <- results_gcn[, .(
  n_folds = uniqueN(fold),
  mean_rmse = mean(rmse),
  mean_mae = mean(mae)
), by = cv_mode]

knitr::kable(summary_gcn[order(cv_mode)], digits = 2)
```

Note what "a batch" costs here versus in Section 5: the MLP's
`dataloader` iterates mini-batches of individual *rows*, and even after
capping training rows at `max_train = 4000` per fold to keep Section 6
fast, the GCN's mini-batches -- individual *days*, at most 365 of them,
each one a single matrix multiply against a 69x69 adjacency shared
across the whole dataset -- need no such subsampling to run quickly.
Refitting all `r sum(sapply(cv_indices, max))` folds across all four
schemes here takes well under a minute despite using more than five
times the epoch count (80 vs. 15), because the number of gradient steps
scales with the number of *days*, not the number of station-day *rows*.

## 13. MLP vs. GCN, side by side

```{r mlp-vs-gcn}
comparison <- rbindlist(list(
  results[, .(cv_mode, rmse, mae, model = "mlp")],
  results_gcn[, .(cv_mode, rmse, mae, model = "gcn")]
))

comparison_summary <- comparison[, .(mean_rmse = mean(rmse)), by = .(cv_mode, model)]
knitr::kable(
  dcast(comparison_summary, cv_mode ~ model, value.var = "mean_rmse")[order(cv_mode)],
  digits = 2
)
```

```{r mlp-vs-gcn-plot, eval = has_deps && requireNamespace("plotly", quietly = TRUE)}
plotly::plot_ly(
  data = comparison,
  x = ~cv_mode, y = ~rmse, color = ~model,
  type = "box"
) |>
  plotly::layout(
    yaxis = list(title = "Fold RMSE (PM10, ug/m3)"),
    xaxis = list(title = "CV scheme"),
    boxmode = "group"
  )
```

## Overall takeaways

* A GNN is not a drop-in improvement -- the naive shared-weight GCN
  layer described in Section 9 was *worse* than the plain row-level MLP
  on this data. The graph structure only helped once the architecture
  (separate self/neighbor weights) let the model keep a station's own
  signal instead of averaging it away with eight neighbors.
* In the run captured above, the GCN and MLP land within noise of each
  other on every scheme -- neither model is a consistent winner, even on
  `lblo`, where a held-out station's graph neighbors are exactly the
  information the MLP does not have direct access to. The likely reason:
  the graph layers here only aggregate the *same* static, exogenous
  features (coordinates, altitude, calendar day) the MLP already
  receives per row -- a neighbor's location does not tell the model much
  that the held-out station's own location did not. A graph is only as
  useful as what flows over its edges; passing neighbors' *observed PM10*
  (e.g. a lagged or same-day value where available) as a node feature,
  not just their static covariates, is the natural next step and would
  give the GCN information genuinely unavailable to the row-level MLP.
* The two models need different data shapes for the *same*
  `stcvlite`-generated fold assignments: the MLP consumes
  `rsample::analysis()`/`assessment()` tables row by row, while the GCN
  needs the row-level `cv_indices` vector reshaped into a
  `(day, station)` mask. `stcvlite`'s CV indices are just integer fold
  labels aligned to your original rows -- reshaping or resampling them to
  fit a given architecture is entirely up to the modeling code, not
  something `stcvlite` needs to know about.
* Mini-batch granularity matters as much as model choice for training
  cost: batching by row (Section 5) versus batching by day over a shared
  graph (Section 12) changed wall-clock training time far more than
  switching from xgboost/BART to a neural network did in the companion
  vignette.
